---
title: 'Reproducible Research: Peer Assessment 1'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r,message=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(Hmisc)

### Read in data 
activity <- read.csv("activity.csv" )
# make dates in POSIXtl form in order to use its methods
#activity$date <-as.POSIXlt(activity$date)
# return dates to Date class
activity$date <- as.Date(activity$date)
# create a new column with day names
activity$day<-factor(weekdays(activity$date))
# use the names to create a factor variable 2 levels:weekdays/weekends
activity$daytype<-factor(activity$day%in% c("Saturday","Sunday"),labels=c('weekday', 'weekend'))

#Quick look 
summary(activity)#NA's   :2304   
```
## What is mean total number of steps taken per day?
```{r,message=FALSE}
## with NAs not removed 
daytotals <- tapply(activity$steps, activity$date, sum)
#histogram of daily totals
hist(daytotals, breaks = 61)  # 61 days 
summary(daytotals) ##median: 10760    mean: 10770
#Using the mean function needs NAs removed
mean(daytotals,na.rm=T) #but with the 8 days removed mean: 10766.19
median(daytotals,na.rm=T) ## median : 10765

# or with dplyr
# You can use data frame or make a data table
# activitydf<-tbl_df(activity)
## daily sum ,NAs included
dsum <-summarise(group_by(activity,date), s = sum(steps))
summary(dsum)
plot(dsum,type="l")##NAs producing gaps in the plot
hist(dsum$s,breaks=17568/1000,main="Histogram of Daily Sums Frequency")

##daily mean, NAs included
dmean <-summarise(group_by(activity,date),m=mean(steps))
summary(dmean)
plot(dmean,type="l")##NAs producing gaps

## Examining by interval (288 intervals per day)
###by interval sum,NAs removed 
isum <-summarise(group_by(activity,interval), s = sum(steps,na.rm=T))
summary(isum)
plot(isum,type="l",ylab="Step sum",main="Sum per interval across all days")
hist(isum$s,breaks=17568/1000,main="Histogram of per Interval Sums Frequency")
```

## What is the average daily activity pattern?
## 1.Make a time series plot of the 5-minute interval(x-axis)
## and the average number of steps taken,averaged across all days (y-axis)
```{r,message=FALSE}
##by interval mean daily steps,NAs removed
imean<-summarise(group_by(activity,interval),m=mean(steps,na.rm=T))
summary(imean)
plot(imean,type="l",xlab="Intervals",ylab="mean steps",main="Weekly mean by interval")
hist(imean$m,breaks=17568/1000,main="Histogram of per Interval Means Frequency")
```

## What is the average daily activity pattern?
```{r,message=FALSE}
#we can plot a Breakdown of steps by day of the week(by name)

meandaytotal<-aggregate(activity$steps, by=list(interval = activity$interval,day = activity$day), FUN=mean,na.rm=T)

library(ggplot2)
ggplot(data = meandaytotal, aes(interval, x, group=day)) +
  geom_line() +
  facet_grid(day ~. ) +
  xlab("Time Interval") +
  ylab("Mean Number of Steps") +
  ggtitle("Mean Number of Steps versus Time Interval")
##qplot 
qplot(meandaytotal$interval,meandaytotal$x,colour=meandaytotal$day)
```

## 2.Which 5-minute interval, on average across all the days in the dataset,contains the maximum number of steps?
```{r,message=FALSE}
## biggest value of mean steps per 5 minute interval 
which.max((isum$s))#result is interval 104
##get the value of the 104 interval
isum[104,]
#gives the value of  10927(most steps/interval across all days)
#Or we can examine the averages across all days ,and find that the
#same interval(104)has the highest average ,with 206.16
which.max((imean$m))
imean[which.max(imean$m),]
```

## Imputing missing values
```{r,message=FALSE}
# NA substituted with mean values
clean<-activity
clean$steps  <- with(activity, impute(steps, mean))
summary(clean)#reports all details

##Repeating the analysis from the start for imputed set
## with NAs not removed 
daytotals <- tapply(clean$steps, clean$date, sum)
#histogram of daily totals
hist(daytotals, breaks = 61)  # 61 days 
summary(daytotals) ##median: 10770    mean: 10770
#Using the mean function needs NAs removed
mean(daytotals,na.rm=T) #but with the 8 days removed mean: 10766.19
median(daytotals,na.rm=T) ## median : 10765

# or with dplyr
# You can use data frame or make a data table
# activitydf<-tbl_df(activity)
## daily sum ,NAs included
dsum <-summarise(group_by(clean,date), s = sum(steps))
summary(dsum)
plot(dsum,type="l")##No gaps in the plot
hist(dsum$s,breaks=17568/1000,main="Histogram of Daily Sums Frequency")

##daily mean, NAs included
dmean <-summarise(group_by(clean,date),m=mean(steps))
summary(dmean)
plot(dmean,type="l")##No gaps

## Examining by interval (288 intervals per day)
###by interval sum,NAs removed 
isum <-summarise(group_by(clean,interval), s = sum(steps,na.rm=T))
summary(isum)
plot(isum,type="l",ylab="Step sum",main="Sum per interval across all days")
hist(isum$s,breaks=17568/1000,main="Histogram of per Interval Sums Frequency")

## What is the average daily activity pattern?
## 1.Make a time series plot of the 5-minute interval(x-axis)
## and the average number of steps taken,averaged across all days (y-axis)

##by interval mean daily steps,NAs removed
imean<-summarise(group_by(clean,interval),m=mean(steps,na.rm=T))
summary(imean)
plot(imean,type="l",xlab="Intervals",ylab="mean steps",main="Weekly mean by interval")
hist(imean$m,breaks=17568/1000,main="Histogram of per Interval Means Frequency")

## What is the average daily activity pattern?
#we can plot a Breakdown of steps by day of the week(by name)

meandaytotal<-aggregate(clean$steps, by=list(interval = clean$interval,day = clean$day), FUN=mean,na.rm=T)

library(ggplot2)
ggplot(data = meandaytotal, aes(interval, x, group=day)) +
  geom_line() +
  facet_grid(day ~. ) +
  xlab("Time Interval") +
  ylab("Mean Number of Steps") +
  ggtitle("Mean Number of Steps versus Time Interval")

##qplot 
qplot(meandaytotal$interval,meandaytotal$x,colour=meandaytotal$day)
```

##Are there di???erent activity patterns between weekdays and weekends?
```{r,message=FALSE}

qplot(clean$interval,clean$steps,colour=clean$daytype)

daytype<-aggregate(clean$steps, by=list(interval = clean$interval,daytype = clean$daytype), FUN=mean,na.rm=T)
p <- ggplot(daytype,aes(x=interval,y=x))
p <- p +geom_line( col= "blue")+facet_grid(daytype~.)
p <- p + labs(title="Average steps per interval by daytype")
p

## With smoothers,we can see clearly the different activity pattern
ggplot(daytype, aes(interval, x, group=daytype)) +
  geom_line() +
  facet_grid(daytype ~ .) +
  geom_point(col='blue', pch=19, cex=0.8, alpha=0.9) +
  geom_rug() +
  ylab("Mean Number of Steps") +
  ggtitle("Mean Number of Steps versus Time Interval")+
  stat_smooth(se=FALSE, method='loess', col='indianred1', lwd=1.2) +
  theme_bw()

```
